---
title: "Analysis of Metabat MAGs from the Barre Woods Filter Metagenomes"
author: "Jeffrey Blanchard"
date: "12/11/2021"
output:
  html_document:
    toc: true
    toc_depth: 4
    theme: united
    highlight: tango
---

# This report is an analysis of the Barre Woods filter metagenomes. Assembly, binning, quality control and classification was done on KBase
(https://narrative.kbase.us/narrative/105147). The CheckM and GTDB files were downloaded to this project folder. I also made used of tables derived from analysis of the metabat bins in VSorter2 which calculated the bin size.

## Load Libraries

```{r}
# Load libraries
library(tidyverse)
library(DT)
```

## Import Data

### Import CheckM and GTDB files. Merge them into a single data frame (tibble)

```{r}
C1_metabat_checkM <- read_tsv("C1/C1_metabat/CheckM_summary_table.tsv")
C1_metabat_gtdb <- read_csv("C1/C1_metabat/gtdb.csv")

# out taxa categories in separate columns
# change bin names in gtdb to remove .fa
C1_metabat_gtdb <- C1_metabat_gtdb %>% 
  mutate_at("User Genome", str_replace, ".fa", "") %>% 
  rename(`Bin Name` = `User Genome`) %>% 
  mutate_at("Classification", str_replace, "d__", "") %>% 
  mutate_at("Classification", str_replace, "p__", "") %>% 
  mutate_at("Classification", str_replace, "c__", "") %>% 
  mutate_at("Classification", str_replace, "o__", "") %>% 
  mutate_at("Classification", str_replace, "f__", "") %>% 
  mutate_at("Classification", str_replace, "g__", "") %>% 
  mutate_at("Classification", str_replace, "s__", "") %>% 
  separate(Classification, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), ";")

# merge columns and add column name 
C1_metabat_merge <- left_join(C1_metabat_checkM,
    C1_metabat_gtdb, by = c("Bin Name")) %>% 
    mutate(Metagenome = "C1") %>% 
    relocate(Metagenome)
```


```{r}
C2_metabat_checkM <- read_tsv("C2/C2_metabat/CheckM_summary_table.tsv")
C2_metabat_gtdb <- read_csv("C2/C2_metabat/gtdb.csv")

# out taxa categories in separate columns
# change bin names in gtdb to remove .fa
C2_metabat_gtdb <- C2_metabat_gtdb %>% 
  mutate_at("User Genome", str_replace, ".fa", "") %>% 
  rename(`Bin Name` = `User Genome`) %>% 
  mutate_at("Classification", str_replace, "d__", "") %>% 
  mutate_at("Classification", str_replace, "p__", "") %>% 
  mutate_at("Classification", str_replace, "c__", "") %>% 
  mutate_at("Classification", str_replace, "o__", "") %>% 
  mutate_at("Classification", str_replace, "f__", "") %>% 
  mutate_at("Classification", str_replace, "g__", "") %>% 
  mutate_at("Classification", str_replace, "s__", "") %>% 
  separate(Classification, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), ";")

# merge columns and add column name 
C2_metabat_merge <- left_join(C2_metabat_checkM,
    C2_metabat_gtdb, by = c("Bin Name")) %>% 
    mutate(Metagenome = "C2") %>% 
    relocate(Metagenome)
```

```{r}
B1_metabat_checkM <- read_tsv("B1/B1_metabat/CheckM_summary_table.tsv")
B1_metabat_gtdb <- read_csv("B1/B1_metabat/gtdb.csv")

# out taxa categories in separate columns
# change bin names in gtdb to remove .fa
B1_metabat_gtdb <- B1_metabat_gtdb %>% 
  mutate_at("User Genome", str_replace, ".fa", "") %>% 
  rename(`Bin Name` = `User Genome`) %>% 
  mutate_at("Classification", str_replace, "d__", "") %>% 
  mutate_at("Classification", str_replace, "p__", "") %>% 
  mutate_at("Classification", str_replace, "c__", "") %>% 
  mutate_at("Classification", str_replace, "o__", "") %>% 
  mutate_at("Classification", str_replace, "f__", "") %>% 
  mutate_at("Classification", str_replace, "g__", "") %>% 
  mutate_at("Classification", str_replace, "s__", "") %>% 
  separate(Classification, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), ";") %>% 
  mutate(`RED Value` = as.character(`RED Value`))

# merge columns and add column name 
B1_metabat_merge <- left_join(B1_metabat_checkM,
    B1_metabat_gtdb, by = c("Bin Name")) %>% 
    mutate(Metagenome = "B1") %>% 
    relocate(Metagenome)
```

```{r}
B2_metabat_checkM <- read_tsv("B2/B2_metabat/CheckM_summary_table.tsv")
B2_metabat_gtdb <- read_csv("B2/B2_metabat/gtdb.csv")

# out taxa categories in separate columns
# change bin names in gtdb to remove .fa
B2_metabat_gtdb <- B2_metabat_gtdb %>% 
  mutate_at("User Genome", str_replace, ".fa", "") %>% 
  rename(`Bin Name` = `User Genome`) %>% 
  mutate_at("Classification", str_replace, "d__", "") %>% 
  mutate_at("Classification", str_replace, "p__", "") %>% 
  mutate_at("Classification", str_replace, "c__", "") %>% 
  mutate_at("Classification", str_replace, "o__", "") %>% 
  mutate_at("Classification", str_replace, "f__", "") %>% 
  mutate_at("Classification", str_replace, "g__", "") %>% 
  mutate_at("Classification", str_replace, "s__", "") %>% 
  separate(Classification, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), ";")

# merge columns and add column name 
B2_metabat_merge <- left_join(B2_metabat_checkM,
    B2_metabat_gtdb, by = c("Bin Name")) %>% 
    mutate(Metagenome = "B2") %>% 
    relocate(Metagenome)
```

```{r}
A1_metabat_checkM <- read_tsv("A1/A1_metabat/CheckM_summary_table.tsv")
A1_metabat_gtdb <- read_csv("A1/A1_metabat/gtdb.csv")

# out taxa categories in separate columns
# change bin names in gtdb to remove .fa
A1_metabat_gtdb <- A1_metabat_gtdb %>% 
  mutate_at("User Genome", str_replace, ".fa", "") %>% 
  rename(`Bin Name` = `User Genome`) %>% 
  mutate_at("Classification", str_replace, "d__", "") %>% 
  mutate_at("Classification", str_replace, "p__", "") %>% 
  mutate_at("Classification", str_replace, "c__", "") %>% 
  mutate_at("Classification", str_replace, "o__", "") %>% 
  mutate_at("Classification", str_replace, "f__", "") %>% 
  mutate_at("Classification", str_replace, "g__", "") %>% 
  mutate_at("Classification", str_replace, "s__", "") %>% 
  separate(Classification, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), ";") %>% 
  mutate(`RED Value` = as.character(`RED Value`))

# merge columns and add column name 
A1_metabat_merge <- left_join(A1_metabat_checkM,
    A1_metabat_gtdb, by = c("Bin Name")) %>% 
    mutate(Metagenome = "A1") %>% 
    relocate(Metagenome)
```


```{r}
A3_metabat_checkM <- read_tsv("A3/A3_metabat/CheckM_summary_table.tsv")
A3_metabat_gtdb <- read_csv("A3/A3_metabat/gtdb.csv")

# out taxa categories in separate columns
# change bin names in gtdb to remove .fa
A3_metabat_gtdb <- A3_metabat_gtdb %>% 
  mutate_at("User Genome", str_replace, ".fa", "") %>% 
  rename(`Bin Name` = `User Genome`) %>% 
  mutate_at("Classification", str_replace, "d__", "") %>% 
  mutate_at("Classification", str_replace, "p__", "") %>% 
  mutate_at("Classification", str_replace, "c__", "") %>% 
  mutate_at("Classification", str_replace, "o__", "") %>% 
  mutate_at("Classification", str_replace, "f__", "") %>% 
  mutate_at("Classification", str_replace, "g__", "") %>% 
  mutate_at("Classification", str_replace, "s__", "") %>% 
  separate(Classification, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), ";")

# merge columns and add column name 
A3_metabat_merge <- left_join(A3_metabat_checkM,
    A3_metabat_gtdb, by = c("Bin Name")) %>% 
    mutate(Metagenome = "A3") %>% 
    relocate(Metagenome)
```

```{r}
# merge tables
all_metabat <- bind_rows(C1_metabat_merge, C2_metabat_merge, B1_metabat_merge, B2_metabat_merge, A1_metabat_merge, A3_metabat_merge)
```


### Import bin (MAG) size (bp) from Vsorter2 data (downloaded from Unity)
```{r}

A1_virsorter_final_calls <- read_tsv("/home/jlb/jlb@umass.edu/GoEcology/Research/GoEcology/2020_NEB_filter_metagenomes/git/filter_metagenomes/virsorter_results/filter_A1_final_calls.tsv") %>% 
  rename(`Bin Name` = `bin`) %>% 
  mutate(Metagenome = "A1") %>% 
  mutate(Pore = "1.2um") %>% 
  relocate(Metagenome)

A3_virsorter_final_calls <- read_tsv("/home/jlb/jlb@umass.edu/GoEcology/Research/GoEcology/2020_NEB_filter_metagenomes/git/filter_metagenomes/virsorter_results/filter_A3_final_calls.tsv") %>% 
  rename(`Bin Name` = `bin`) %>% 
  mutate(Metagenome = "A3") %>% 
  mutate(Pore = "0.2um") %>% 
  relocate(Metagenome)

B1_virsorter_final_calls <- read_tsv("/home/jlb/jlb@umass.edu/GoEcology/Research/GoEcology/2020_NEB_filter_metagenomes/git/filter_metagenomes/virsorter_results/filter_B1_final_calls.tsv") %>% 
  rename(`Bin Name` = `bin`) %>% 
  mutate(Metagenome = "B1") %>% 
  mutate(Pore = "1.2um") %>% 
  relocate(Metagenome)

B2_virsorter_final_calls <- read_tsv("/home/jlb/jlb@umass.edu/GoEcology/Research/GoEcology/2020_NEB_filter_metagenomes/git/filter_metagenomes/virsorter_results/filter_B2_final_calls.tsv") %>% 
  rename(`Bin Name` = `bin`) %>% 
  mutate(Metagenome = "B2") %>% 
  mutate(Pore = "0.2um") %>% 
  relocate(Metagenome)

C1_virsorter_final_calls <- read_tsv("/home/jlb/jlb@umass.edu/GoEcology/Research/GoEcology/2020_NEB_filter_metagenomes/git/filter_metagenomes/virsorter_results/filter_C1_final_calls.tsv") %>% 
  rename(`Bin Name` = `bin`) %>% 
  mutate(Metagenome = "C1") %>% 
  mutate(Pore = "1.2um") %>% 
  relocate(Metagenome)

C2_virsorter_final_calls <- read_tsv("/home/jlb/jlb@umass.edu/GoEcology/Research/GoEcology/2020_NEB_filter_metagenomes/git/filter_metagenomes/virsorter_results/filter_C2_final_calls.tsv") %>% 
  rename(`Bin Name` = `bin`) %>% 
  mutate(Metagenome = "C2") %>% 
  mutate(Pore = "0.2um") %>% 
  relocate(Metagenome)

```

```{r}
#merge individual metagenomes into a single table
bin_size <- bind_rows(A1_virsorter_final_calls, A3_virsorter_final_calls, B1_virsorter_final_calls, B2_virsorter_final_calls, C1_virsorter_final_calls, C2_virsorter_final_calls)
```

### Join CheckM, GTDB and MAG size tables

```{r}
all_metabat_bin_size <- left_join(all_metabat,
    bin_size, by = c("Metagenome", "Bin Name")) %>% 
    select(Metagenome, `Bin Name`, Completeness, Contamination, Domain, Phylum, Class, Order, Family, Genus, Species, BinContigLenSum, gcMean)
  
```

```{r, eval=FALSE}
# write summary file
write_tsv(all_metabat_bin_size, "all_metabat_bin_size.tsv") 

```

## Data analysis

### Filter low quality MAGs

```{r}
all_metabat_bin_size_90_10 <- all_metabat_bin_size %>%
  filter(Completeness > 89.5) %>% 
  filter(Contamination < 10.4) 
write_tsv(all_metabat_bin_size_90_10, "all_metabat_bin_size_90_10")
```
```{r}
all_metabat_bin_size_50_10 <- all_metabat_bin_size %>%
  filter(Completeness > 49.5) %>% 
  filter(Contamination < 10.4) 
write_tsv(all_metabat_bin_size_50_10, "all_metabat_bin_size_50_10")
```

#### Table of Medium and High Quality MAGs

```{r}
datatable(all_metabat_bin_size_50_10 %>% 
  group_by(Phylum) %>% 
  summarise(n = n()) %>%
  mutate(freq = n / sum(n)) %>% 
  arrange(-n)
)
```

#### Table of High Quality MAGs

```{r}
datatable(all_metabat_bin_size_90_10 %>% 
  group_by(Phylum) %>% 
  summarise(n = n()) %>%
  mutate(freq = n / sum(n)) %>% 
  arrange(-n)
)
```


#### Phyla

```{r, echo=FALSE}
all_metabat_50_10 %>% 
  count(Phylum, sort = TRUE) %>% 
  ggplot(aes(x = "", y = n, fill = reorder(Phylum, -n))) + 
    geom_bar(width = 1, stat = "identity") +
    coord_polar(theta = "y", start=0) +
 #   geom_text(aes(label = paste0(n)), position = position_stack(vjust=0.5)) +
    labs(fill="Phylum", 
       x=NULL, 
       y=NULL, 
       title="Phyla in filter MAGs")
```
#### Class

```{r, echo=FALSE}
all_metabat_50_10 %>% 
  count(Class, sort = TRUE) %>% 
  ggplot(aes(x = "", y = n, fill = reorder(Class, -n))) + 
    geom_bar(width = 1, stat = "identity") +
    coord_polar(theta = "y", start=0) +
 #   geom_text(aes(label = paste0(n)), position = position_stack(vjust=0.5)) +
    labs(fill="Class", 
       x=NULL, 
       y=NULL, 
       title="Class in filter MAGs")
```


#### Order

```{r, echo=FALSE}
all_metabat_50_10 %>% 
  count(Order, sort = TRUE) %>% 
  ggplot(aes(x = "", y = n, fill = reorder(Order, -n))) + 
    geom_bar(width = 1, stat = "identity") +
    coord_polar(theta = "y", start=0) +
 #   geom_text(aes(label = paste0(n)), position = position_stack(vjust=0.5)) +
    labs(fill="Order", 
       x=NULL, 
       y=NULL, 
       title="Order in filter MAGs")
```


#### Phyla bar chart


```{r, echo=FALSE}
all_metabat_50_10 %>% 
  count(Phylum, sort = TRUE) %>% 
  ggplot(aes(x=reorder(Phylum, n), y = n)) +
  geom_col(colour = "maroon", fill = "maroon") +
  coord_flip()+
  ggtitle("Number of MAGs for each Phylum") +
  ylab("Number of MAGs") + 
  xlab("Class")
```


#### Class bar chart


```{r, echo=FALSE}
all_metabat_50_10 %>% 
  count(Class, sort = TRUE) %>% 
  ggplot(aes(x=reorder(Class, n), y = n)) +
  geom_col(colour = "maroon", fill = "maroon") +
  coord_flip()+
  ggtitle("Number of MAGs for each Class") +
  ylab("Number of MAGs") + 
  theme(text = element_text(size = 20, color="black")) +
  xlab("Class")
```



```{r}
all_metabat_virsorter_50_10 <-all_metabat_virsorter %>%
  filter(Completeness > 50) %>% 
  filter(Contamination < 10) 
```



```{r}
all_metabat_virsorter_50_10 %>% 
ggplot(aes(x = Class, y = BinContigLenSum, color = Domain)) + 
  geom_point() +
  ggtitle("Number of MAGs for each Class") +
  ylab("Number of MAGs") + 
  xlab("Class") +
  coord_flip()
#  theme(axis.text.x = element_text(angle = 90)) 
```

```{r}
all_metabat_virsorter_50_10 %>% 
ggplot(aes(x = reorder(Class, BinContigLenSum, na.rm = TRUE), y = BinContigLenSum, color = Domain)) + 
  geom_boxplot() +
  ggtitle("Number of MAGs for each Class") +
  ylab("Number of MAGs") + 
  xlab("Class") +
  coord_flip()
#  theme(axis.text.x = element_text(angle = 90)) 
```


```{r}
all_metabat_virsorter_50_10 %>% 
ggplot(aes(x = reorder(Class, BinContigLenSum, na.rm = TRUE), y = BinContigLenSum, color = Domain)) + 
  geom_point() +
  ggtitle("Genome size of MAGs for each Class") +
  ylab("Genome size") + 
  xlab("Class") +
  coord_flip() +
  theme(text = element_text(size = 20, color="black"))
#  theme(axis.text.x = element_text(angle = 90)) 
```

## Histogram

```{r}
all_metabat_virsorter_50_10 %>% 
ggplot(aes(x = BinContigLenSum)) + 
  geom_histogram(colour = "black", fill = "maroon", binwidth=500000) +
  ggtitle("Genome size of MAGs for each Class") +
  ylab("Genome size") + 
  theme(text = element_text(size = 20, color="black"))
#  theme(axis.text.x = element_text(angle = 90)) 
```

ggplot(df, aes(x=weight, color=sex)) +
  geom_histogram(fill="white")
  
  
```{r}
all_metabat_virsorter_50_10 %>% 
ggplot(aes(x = reorder(Class, BinContigLenSum, na.rm = TRUE), y = BinContigLenSum, color = Pore)) + 
  geom_point() +
  ggtitle("Genome size of MAGs for each Class") +
  ylab("Number of MAGs") + 
  xlab("Class") +
  coord_flip()
#  theme(axis.text.x = element_text(angle = 90)) 
```

```{r}
all_metabat_virsorter_50_10 %>% 
ggplot(aes(x = reorder(Class, BinContigLenSum, na.rm = TRUE), y = BinContigLenSum, color = Metagenome)) + 
  geom_point() +
  ggtitle("Genome size of MAGs for each Class") +
  ylab("Number of MAGs") + 
  xlab("Class") +
  coord_flip()
#  theme(axis.text.x = element_text(angle = 90)) 
```


```{r}
all_metabat_virsorter_50_10 %>% 
ggplot(aes(x = reorder(Phylum, BinContigLenSum, na.rm = TRUE), y = BinContigLenSum, color = Domain)) + 
  geom_point() +
  ggtitle("Number of MAGs for each Phylum") +
  ylab("Number of MAGs") + 
  xlab("Phylum") +
  coord_flip()
#  theme(axis.text.x = element_text(angle = 90)) 
```



```{r}
all_metabat_virsorter_50_10 %>% 
ggplot(aes(x = reorder(Order, BinContigLenSum, na.rm = TRUE), y = BinContigLenSum, color = Domain)) + 
  geom_point() +
  ggtitle("Number of MAGs for each Order") +
  ylab("Number of MAGs") + 
  xlab("Order") +
  coord_flip()
#  theme(axis.text.x = element_text(angle = 90)) 
```


```{r}
write_tsv(all_metabat_virsorter, "all_metabat_virsorter.tsv")
write_tsv(all_metabat_virsorter_50_10, "all_metabat_virsorter_50_10")
```

# Phages

```{r}
all_metabat_virsorter_phage <- all_metabat_virsorter %>% 
  filter(bin_high_group == "dsDNAphage")
```


```{r}
all_metabat_virsorter_phage %>% 
ggplot(aes(x = bin_high_group, y = BinContigLenSum)) + 
  geom_point(colour = "maroon", fill = "maroon") +
  geom_jitter(colour = "maroon", fill = "maroon") +
  ggtitle("Genome size of dsDNA phage MAGs") +
  ylab("Number of MAGs") + 
  xlab("dsDNA phage MAGs") +
  coord_flip()
#  theme(axis.text.x = element_text(angle = 90)) 
```



```{r}
all_metabat_virsorter_phage %>% 
ggplot(aes(x = bin_high_group, y = BinContigLenSum)) + 
  geom_jitter(colour = "maroon", fill = "maroon") +
  ggtitle("Genome size of dsDNA phage MAGs") +
  ylab("Genome size") + 
  xlab("dsDNA phage MAGs") +
  theme(text = element_text(size = 20, color="black")) +
  coord_flip()
#  theme(axis.text.x = element_text(angle = 90)) 
```

```{r}
all_metabat_virsorter_phage %>% 
ggplot(aes(x = bin_high_group, y = BinContigLenSum)) + 
  geom_point() +
  coord_flip()
#  theme(axis.text.x = element_text(angle = 90)) 
```

