---
title: "Analysis of Metabat MAGs from the Barre Woods Filter Metagenomes"
author: "Jeffrey Blanchard"
date: "9/16/2022"
output:
  html_document:
    toc: true
    toc_depth: 4
    theme: united
    highlight: tango
---

# This report is an analysis of the Barre Woods filter metagenomes. Assembly, binning, quality control and classification was done on KBase
(https://narrative.kbase.us/narrative/105147). The CheckM and GTDB files were downloaded to this project folder. I also made used of tables derived from analysis of the metabat bins in VSorter2 which calculated the bin size.

## Load Libraries

```{r}
# Load libraries
library(tidyverse)
library(DT)
```
### Load Data

```{r}
all_metabat_bin_size_50_10 <- read_tsv("all_metabat_bin_size_50_10.tsv",
col_types = 
  cols(
    Domain = col_factor(),
    Phylum = col_factor(),
    Class = col_factor(),
    Order = col_factor(),
    Family = col_factor(),
    Genus  = col_factor(),
    Species = col_factor(),
    Metagenome  = col_factor()
  ))
```


## Graphs

### Phyla pie chart
```{r}
all_metabat_bin_size_50_10 %>% 
  count(Phylum, sort = TRUE) %>% 
  ggplot(aes(x = "", y = n, fill = reorder(Phylum, -n))) + 
    geom_bar(width = 1, stat = "identity") +
    coord_polar(theta = "y", start=0) +
 #   geom_text(aes(label = paste0(n)), position = position_stack(vjust=0.5)) +
    labs(fill="Phylum", 
       x=NULL, 
       y=NULL, 
       title="Phyla in filter MAGs")
```

```{r}
all_metabat_bin_size_50_10 %>% 
  count(Phylum, sort = TRUE) %>% 
  ggplot(aes(x=reorder(Phylum, n), y = n)) +
  geom_col(colour = "maroon", fill = "maroon") +
  coord_flip()+
  ggtitle("Number of MAGs for each Phylum") +
  ylab("Number of MAGs") + 
  xlab("Phylum") 
```
```{r}
all_metabat_bin_size_50_10 %>% 
  filter(BinContigLenSum <2000000) %>% 
  count(Phylum, sort = TRUE) %>% 
  ggplot(aes(x=reorder(Phylum, n), y = n)) +
  geom_col(colour = "maroon", fill = "maroon") +
  coord_flip()+
  ggtitle("Number of MAGs for each Phylum") +
  ylab("Number of MAGs < 2MB") + 
  xlab("Phylum") 
```


```{r}
all_metabat_bin_size_50_10 %>% 
  count(Metagenome, Phylum, sort = TRUE) %>% 
  ggplot(aes(x=reorder(Phylum, n), y = n, fill = Metagenome)) +
  geom_col() +
  coord_flip()+
  ggtitle("Number of MAGs for each Phylum") +
  ylab("Number of MAGs") + 
  xlab("Phylum") 
```

### 

```{r, fig.height=8}
all_metabat_bin_size_50_10 %>% 
  count(Metagenome, Phylum, sort = TRUE) %>% 
  ggplot(aes(x=reorder(Phylum, n), y = n, fill = Metagenome)) +
  geom_col() +
  coord_flip()+
  ggtitle("Number of MAGs for each Phylum") +
  ylab("Number of MAGs") + 
  xlab("Phylum") +
  facet_wrap(~Metagenome)
```



```{r, fig.height=8}
all_metabat_bin_size_50_10 %>% 
  filter(BinContigLenSum <2000000) %>% 
  count(Metagenome, Phylum, sort = TRUE) %>% 
  ggplot(aes(x=reorder(Phylum, n), y = n, fill = Metagenome)) +
  geom_col() +
  coord_flip()+
  ggtitle("Number of MAGs for each Phylum") +
  ylab("Number of MAGs") + 
  xlab("Phylum") +
  facet_wrap(~Metagenome)
```




### Class bar chart

```{r, fig.height=10}
all_metabat_bin_size_50_10 %>% 
  count(Class, sort = TRUE) %>% 
  ggplot(aes(x=reorder(Class, n), y = n)) +
  geom_col(colour = "maroon", fill = "maroon") +
  coord_flip()+
  ggtitle("Number of MAGs for each Class") +
  ylab("Number of MAGs") + 
  xlab("Class")
```

### Order bar chart

```{r, fig.height=10}
all_metabat_bin_size_50_10 %>% 
  count(Order, sort = TRUE) %>% 
  ggplot(aes(x=reorder(Order, n), y = n)) +
  geom_col(colour = "maroon", fill = "maroon") +
  coord_flip()+
  ggtitle("Number of MAGs for each Order") +
  ylab("Number of MAGs") + 
  xlab("Order")
```



### Histogram of MAG size distribution

```{r}
all_metabat_bin_size_50_10 %>% 
ggplot(aes(x = BinContigLenSum)) + 
  geom_histogram(colour = "black", fill = "maroon", binwidth=500000) +
  ggtitle("Genome size of MAGs for each Class") +
  ylab("Genome size") + 
  xlab("MAG size (bp)") +
  theme(text = element_text(size = 20, color="black"))
```

```{r}
all_metabat_bin_size_50_10 %>% 
ggplot(aes(x = BinContigLenSum, fill = Metagenome)) + 
  geom_histogram(colour = "black", binwidth=500000) +
  ggtitle("Genome size of MAGs for each Class") +
  ylab("Genome size") + 
  xlab("MAG size (bp)") +
  theme(text = element_text(size = 20, color="black"))
```

```{r}
all_metabat_bin_size_50_10 %>% 
ggplot(aes(x = BinContigLenSum, fill = Metagenome)) + 
  geom_histogram(colour = "black", binwidth=500000) +
  ggtitle("Genome size of MAGs for each Class") +
  ylab("Genome size") + 
  xlab("MAG size (bp)") +
  facet_wrap(~Metagenome) +
  theme(text = element_text(size = 8, color="black"))
```


### xyplot of MAG Size for each class
```{r}
all_metabat_bin_size_50_10 %>% 
ggplot(aes(x = reorder(Class, BinContigLenSum, na.rm = TRUE), y = BinContigLenSum, color = Domain)) + 
  geom_point() +
  ggtitle("Genome size of MAGs for each Class") +
  ylab("MAG size (bp)") + 
  xlab("Class") +
  coord_flip()
```
### xyplot of MAG Size for each class with jitter
```{r, fig.height = 8}
all_metabat_bin_size_50_10 %>% 
ggplot(aes(x = reorder(Class, BinContigLenSum, na.rm = TRUE), y = BinContigLenSum, color = Domain)) + 
  geom_jitter() +
  ggtitle("Genome size of MAGs for each Class") +
  ylab("MAG size (bp)") + 
  xlab("Class") +
  coord_flip()
```

### Boxplot of MAG size for each class

```{r}
all_metabat_bin_size_50_10 %>% 
ggplot(aes(x = reorder(Class, BinContigLenSum, na.rm = TRUE), y = BinContigLenSum, color = Domain)) + 
  geom_boxplot() +
  ggtitle("Number of MAGs for each Class") +
  ylab("MAG size (bp)") + 
  xlab("Class") +
  coord_flip()
```


### xyplot of MAG size for each class with metagenome data set
```{r}
all_metabat_bin_size_50_10 %>% 
ggplot(aes(x = reorder(Class, BinContigLenSum, na.rm = TRUE), y = BinContigLenSum, color = Metagenome)) + 
  geom_point() +
  ggtitle("Genome size of MAGs for each Class") +
  ylab("MAG size (bp)") + 
  xlab("Class") +
  coord_flip()
```

